Appendix to "New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-02-17 jso.qmd
edited sec 6, moved Fig 5 to appendix

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(tourr)
library(PPtreeViz)
library(ferrn)
library(colorspace)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

In addition, to assess the performance of JSO across various PP problems, two different
data shapes, pipe and a sine wave, are considered in various dimensional spaces
using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. A list of combinations of the data shape, dimensions and PP indexes considered are provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

```{r}
#| label: fig-plot-data
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "The sine and pipe structures in the example data, along with index values for these projections."
data("sine1000")
sine1000 <- as_tibble(sine1000)
dp1 <- ggplot(sine1000, aes(x=V5, y=V6)) + geom_point()
data("pipe1000")
pipe1000 <- as_tibble(pipe1000)
dp2 <- ggplot(pipe1000, aes(x=V5, y=V6)) + geom_point()
dp1 + dp2 + plot_layout(ncol=2)
```

@@

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For problems with relatively lower dimensional search spaces (4 and 6 dimensions), small parameter values (20 jellyfish and a maximum of 50 tries) are sufficient for space exploration and thus can already achieve a high success rate. Higher parameter values (i.e. 100 jellyfish and a maximum of 100 tries) enhances exploration of the search space, which is particularly beneficial in higher-dimensional problems (8, 10, and 12 dimensions) where a more complex search is necessary. However, in lower-dimensional problems, the benefit may be less pronounced, as the space is alredy relatively easy to navigate. While increasing both parameters enhances the performance of JSO, it also extends the computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies. As expected, the JSO success rate is higher with higher squintability and/or more jellyfish, and lower when dimension is higher (higher-dimensional optimization is more difficult). Interestingly, the JSO is unaffected by the smoothness of the index function. This is consistent with the way random search algorithms jump from value to value without taking local regularity into account, as opposed to gradient-based optimizers for example. Allowing
for more tries also does not affect success rate significantly. A unit increase in squintability increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As dimension
increases by one the success rate almost halves. Increasing the number of
jellies by 10, increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.

@@

>>>>

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(tourr)
library(PPtreeViz)
library(ferrn)
library(colorspace)
library(broom)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

In addition, to assess the performance of JSO across various PP problems, two different
data shapes, pipe and a sine wave, are considered in various dimensional spaces
using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For problems with relatively lower dimensional search spaces (4 and 6 dimensions), small parameter values (20 jellyfish and a maximum of 50 tries) are sufficient for space exploration and thus can already achieve a high success rate. Higher parameter values (i.e. 100 jellyfish and a maximum of 100 tries) enhances exploration of the search space, which is particularly beneficial in higher-dimensional problems (8, 10, and 12 dimensions) where a more complex search is necessary. However, in lower-dimensional problems, the benefit may be less pronounced, as the space is already relatively easy to navigate. While increasing both parameters enhances the performance of JSO, it also extends the computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies. As expected, the JSO success rate is higher with higher squintability and/or more jellyfish, and lower when dimension is higher (higher-dimensional optimization is more difficult). Interestingly, the JSO is unaffected by the smoothness of the index function. This is consistent with the way random search algorithms jump from value to value without taking local regularity into account, as opposed to gradient-based optimizers for example. Allowing
for more tries also does not affect success rate significantly. A unit increase in squintability increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As dimension
increases by one the success rate almost halves. Increasing the number of
jellies by 10 increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.

@@

```{r}
#| tbl-pos: H
#| label: tbl-joint-output
library(kableExtra)
mod2 <- glm(proportion ~ squintability + d + n_jellies + max_tries, data = sim_df2, family = binomial)
mod3 <- glm(proportion ~ smoothness + d + n_jellies + max_tries, data = sim_df2, family = binomial)
summary_table <- bind_rows(
  glance(mod2) %>% mutate(Model = "Squintability"),
  glance(mod3) %>% mutate(Model = "Smoothness")
) %>%
  select(Model, null.deviance, df.null, logLik, df.residual, AIC, BIC) %>%
  rename(
    `Null Deviance` = null.deviance,
    `DF Null` = df.null,
    `Log-Likelihood` = logLik,
    `DF Residual` = df.residual
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))
summary_table %>%
  kable(caption = "Model summaries for squintability and smoothness separately") %>%
  kable_styling(font_size = 10) %>%
  column_spec(1, bold = TRUE, width = "5em")
```

Squintability and smoothness are likely dependent, making it unnecessary to include both as independent factors in the model. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness.

This study intends to serve as an illustration that the proposed measure effectively captures the key factors influencing the success rate of PP. The interactions between the two variables were not examined in this study due to the limited number of data points, nor was the linearity assumption of the relationship with success rate explicitly tested.

### 2025-02-17 jso.tex
edited sec 6, moved Fig 5 to appendix

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, rightrule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, leftrule=.75mm, toptitle=1mm, colback=white, titlerule=0mm, breakable, opacityback=0, opacitybacktitle=0.6, bottomtitle=1mm, toprule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, coltitle=black]

@@

In addition, to assess the performance of JSO across various PP
problems, two different data shapes, pipe and a sine wave, are
considered in various dimensional spaces using six different PP indexes:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, under varied JSO
hyper-parameters. This results in a total of 76 scenarios, comprising of
30 computed on the pipe data and 46 on the sine-wave data. A list of
combinations of the data shape, dimensions and PP indexes considered are
provided in Table~\ref{tbl-smoothness-squintability}. Again, JSO is run
50 times for each scenario to calculate the success rate for each
hyper-parameter combination.

\includegraphics{jso_files/figure-pdf/fig-plot-data-1.pdf}

\caption{\label{fig-plot-data}The sine and pipe structures in the
example data, along with index values for these projections.}

@@

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For problems
with relatively lower dimensional search spaces (4 and 6 dimensions),
small parameter values (20 jellyfish and a maximum of 50 tries) are
sufficient for space exploration and thus can already achieve a high
success rate. Higher parameter values (i.e.~100 jellyfish and a maximum
of 100 tries) enhances exploration of the search space, which is
particularly beneficial in higher-dimensional problems (8, 10, and 12
dimensions) where a more complex search is necessary. However, in
lower-dimensional problems, the benefit may be less pronounced, as the
space is alredy relatively easy to navigate. While increasing both
parameters enhances the performance of JSO, it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies.
As expected, the JSO success rate is higher with higher squintability
and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more tries also does
not affect success rate significantly. A unit increase in squintability
increases the success rate by 24\%. As dimension increases by one the
success rate almost halves. Increasing the number of jellies by 10,
increases the success rate by 32\%.

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, arc=.35mm, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, rightrule=.15mm, toptitle=1mm, bottomrule=.15mm, breakable, titlerule=0mm, colframe=quarto-callout-note-color-frame]

@@

In addition, to assess the performance of JSO across various PP
problems, two different data shapes, pipe and a sine wave, are
considered in various dimensional spaces using six different PP indexes:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, under varied JSO
hyper-parameters. This results in a total of 76 scenarios, comprising of
30 computed on the pipe data and 46 on the sine-wave data. A list of
combinations of the data shape, dimensions and PP indexes that are
considered in this study is provided in
Table~\ref{tbl-smoothness-squintability}. Again, JSO is run 50 times for
each scenario to calculate the success rate for each hyper-parameter
combination.

@@

The effect of JSO hyper-parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For problems
with relatively lower dimensional search spaces (4 and 6 dimensions),
small parameter values (20 jellyfish and a maximum of 50 tries) are
sufficient for space exploration and thus can already achieve a high
success rate. Higher parameter values (i.e.~100 jellyfish and a maximum
of 100 tries) enhances exploration of the search space, which is
particularly beneficial in higher-dimensional problems (8, 10, and 12
dimensions) where a more complex search is necessary. However, in
lower-dimensional problems, the benefit may be less pronounced, as the
space is already relatively easy to navigate. While increasing both
parameters enhances the performance of JSO, it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies.
As expected, the JSO success rate is higher with higher squintability
and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more tries also does
not affect success rate significantly. A unit increase in squintability
increases the success rate by 24\%. As dimension increases by one the
success rate almost halves. Increasing the number of jellies by 10
increases the success rate by 32\%.

@@

\begingroup\fontsize{10}{12}\selectfont

\begin{longtable}[t]{>{\raggedright\arraybackslash}p{5em}rrrrrr}

\caption{\label{tbl-joint-output}Model summaries for squintability and
smoothness separately}

\tabularnewline

\\
\toprule
Model & Null Deviance & DF Null & Log-Likelihood & DF Residual & AIC & BIC\\
\midrule
\textbf{Squintability} & 49.29 & 75 & -26.44 & 71 & 62.87 & 74.53\\
\textbf{Smoothness} & 49.29 & 75 & -33.23 & 71 & 76.47 & 88.12\\
\bottomrule

\end{longtable}

\endgroup{}

Squintability and smoothness are likely dependent, making it unnecessary
to include both as independent factors in the model. Intuitively, a
highly smooth search space exhibits gradual changes, making it easier to
perceive the overall structure from a distance, thereby increasing the
squint angle. Table~\ref{tbl-joint-output} compare the model fits when
squintability and smoothness are included separately. Given that the
null deviance remains the same across models, the model with
squintability yields a smaller residual deviance and lower AIC and BIC
values, indicating a better fit. This suggests that squintability
captures the relevant variation more effectively than smoothness.

This study intends to serve as an illustration that the proposed measure
effectively captures the key factors influencing the success rate of PP.
The interactions between the two variables were not examined in this
study due to the limited number of data points, nor was the linearity
assumption of the relationship with success rate explicitly tested.

### 2025-02-17 supplementary/index-def.qmd
edited sec 6, moved Fig 5 to appendix

>>>>

# The sine and pipe structures in the example data

```{r}
#| label: fig-plot-data
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "The sine and pipe structures in the example data, along with index values for these projections."
data("sine1000")
sine1000 <- as_tibble(sine1000)
dp1 <- ggplot(sine1000, aes(x=V5, y=V6)) + geom_point()
data("pipe1000")
pipe1000 <- as_tibble(pipe1000)
dp2 <- ggplot(pipe1000, aes(x=V5, y=V6)) + geom_point()
dp1 + dp2 + plot_layout(ncol=2)
```
